---
title: "Class11"
author: "Helena Khalil A16313711"
format: pdf
---


#Section 1, Proportion of G/G in a population 

Downloded a CSV file from Ensemble 
https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077#_

Here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```


Lets look at a different population. I picked the GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(gbr)
```
Find proportion of GIG 
```{r}
round(table(gbr$Genotype..forward.strand.)/ nrow(gbr)*100,2)
```
This varient which is associated with child asthma is more frequent in the GBR population than the MKL population 

Lets now dig into this further 


Section 4 

How many samples do we have? 
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```
Lets make a boxplot 

```{r}
ggplot(expr)+aes(geno, exp, fill=geno)+geom_boxplot(notch=TRUE)
```

