---
title: "Class09"
author: "Helena Khalil #A16313711"
format: pdf
---
```{r}
read.csv("candy-data.txt")
```
```{r}
candy_file <- "candy-data.txt"

candy = read.csv(candy_file, row.names=1)
head(candy)
```


Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```


Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```
I can convert the 1 and 0 values to be TRUE or FAlSE and use that to extract the type of candy I want. For example the chocolate candy...
```{r}
candy[as.logical(candy$chocolate),]
```



Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Sour Patch Kids", ]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```
Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent has a different scale, it is not in the 0-1 scale 

Q7. What do you think a zero and one represent for the candy$chocolate column?

0 is for false for if its a chocolate and 1 is true for if its a chocolate 

Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy)+ 
  aes(winpercent)+ 
  geom_histogram(binwidth=10)
```


Q9. Is the distribution of winpercent values symmetrical?
no it is not symetrical

Q10. Is the center of the distribution above or below 50%?
No it is below 50% distribution 

Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds,"winpercent"]
choc.win
```
```{r}
#Do the same fro fruity 
fruit.inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.inds,"winpercent"]
fruit.win
```

```{r}
mean(choc.win)
```
```{r}
mean(fruit.win)
```

Chocolate is the winner, it has higher winpercent values!! 

Q12. Is this difference statistically significant?
```{r}
t.test(choc.win, fruit.win)
```
Overall Candy Ranking 


Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```

Q14. What are the top 5 all time favorite candy types out of this set?
Reeses's Peanut Butter cups, Reese's Minatures, Twix, KitKat, Snickers 
```{r}
tail(candy[order(candy$winpercent),], n=5)
```

```{r}
mycols <- rep("gray", nrow(candy))
#mycols[2:5] <- "red"
mycols[as.logical(candy$fruity)] <- "red"
mycols[as.logical(candy$chocolate)] <- "brown"
mycols
```

Q15. Make a first barplot of candy ranking based on winpercent values.
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent))+
  geom_col(fill=mycols)
```


Q17. What is the worst ranked chocolate candy?
Sixlets 

-Q18. What is the best ranked fruity candy?
Starbursts

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Reeses's Minatures 


Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


Exploring the Correlation Structure 
```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Fruit and Chocolate 

Q23. Similarly, what two variables are most positively correlated?
Winpercent and Chocolate 


Principal Component Analysis 


```{r}
pca <- prcomp(candy,scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2], col=mycols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#ggplotly(p)
```

```{r}
library(plotly)
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity and Pluribus, These make sense to me because fruity candy comes in the bag with multiple candies. 
